Multi-target detection with application to cryo-electron microscopy

### 2019-03-05 inverse_problems/inverse_problems.tex
sec 4.2

In what follows, we consider autocorrelations of both the observation $y$ and of the signal occurrences in $y$. As per our discussion of extensions, the signal occurrences may be sampled from a discrete set $\{x_1, \ldots, x_K\}$ (as in~\eqref{eq:heterogeneousdistribution}), or from a more general distribution. Accordingly, we define autocorrelations broadly for a random signal $z$ of length $M$. For our purposes, this will be applied both to signal occurrences (of length $L$) and to $y$ (of length $N$).

For a random signal $z\in \R^M$, the autocorrelation of order $q = 1, 2, \ldots$ is given for any integer shifts $\ell_1, \ldots, \ell_{q-1}$ by
\begin{equation}
a_z^q[\ell_1,\ldots,\ell_{q-1}]   = \E_z\left\{\frac{1}{M} \sum_{i=-\infty}^{\infty} z[i]z[i+\ell_1]\cdots z[i+\ell_{q-1}]\right\},
\label{eq:ac_general}
\end{equation}
where the expectation is taken with respect to the distribution of $z$. Indexing out of bounds is zero-padded, that is, $z[i] = 0$ for $i$ out of the range $0, \ldots, M-1$.
Explicitly, the first-, second- and third-order autocorrelations are given by: 
\begin{align}  
a_z^1 & = \E_z\left\{\frac{1}{M} \sum_{i=0}^{M-1} 
z[i]\right\}, \nonumber\\
a_z^2[\ell] & = \E_z\left\{\frac{1}{M} \sum_{i = \max\{0, -\ell\}}^{M-1 + \min\{0, -\ell\}} z[i]z[i+\ell]\right\}, \label{eq:ac_special} \\
a_z^3[\ell_1,\ell_2] & = \E_z\left\{\frac{1}{M} \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{M-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]\right\}. \nonumber
\end{align}
Since  autocorrelations depend only on the differences between indices, they obey the following symmetries: $$a_z^2[\ell] = a_z^2[-\ell],$$ and
$$a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].
$$
%In what follows, $x$ is the random variable corresponding to signal occurrences in $y$.
% In particular, for the model~\eqref{eq:model}, 
%$x=x_k$ with probability $\pi_k$.
In particular, for
$x$ sampled from $\{ x_1, \ldots, x_K \}$ with probabilities $(\pi_1, \ldots, \pi_K)$ as in~\eqref{eq:heterogeneousdistribution}, the autocorrelations of $x$ are given in explicit form in terms of those of the deterministic signals $x_1, \ldots, x_K$ as:
\begin{align}
	a_x^q  & = \sum_{k=1}^{K} \pi_ka_{x_k}^q.
	\label{eq:mixedautocorr}
\end{align}
Explicit expressions of the autocorrelations for the more involved model of  cryo-EM~\eqref{eq:cryoEM} are given in~\cite{bendory2018toward}.

We are given one observation (one realization) of $y$. Thus, we cannot compute the autocorrelations of $y$ exactly as they involve taking an expectation against the distribution of $y$. However, by the law of large numbers, as $N$ grows to infinity with $\gamma$ remaining constant \TODO{We didn't define $\gamma$ for the well-separated model at this point; we should reconsider this}, the empirical autocorrelations of $y$ almost surely (a.s.) converge to the actual (population) autocorrelations of $y$, that is, % (ergodicity)
\begin{align}
\lim_{N \to \infty, \gamma \textrm{ constant}} \frac{1}{N} \sum_{i=-\infty}^{\infty} y[i]y[i+\ell_1]\cdots y[i+\ell_{q-1}] \aseq a_y^q[\ell_1, \ldots, \ell_{q-1}].
\end{align}
\TODO{Why do we need constant $\gamma$?}
This provides a concrete means of estimating the quantities $a_y^ q$. In the remainder of this section, we relate the observables $a_y^q$  to the unknowns $a_x^q$, first under the well-separated model, then under the Poisson model.

@@

Similarly to the homogeneous well-separated model, the observed autocorrelations under the Poisson model identify $x,\gamma$ and $\sigma^2$ uniquely. Furthermore, these quantities can be computed explicitly.

We first note that $\gamma a_x^1$, $\gamma a_x^2[\ell] + \sigma^2 \ell[\ell]$, and $\gamma a_x^3[\ell_1,\ell_2]$ can be computed directly form the observed autocorrelations~\eqref{eq:mean_micrograph2},~\eqref{eq:ac2_micrograph2} and~\eqref{eq:ac3_micrograph2}.
 Indeed, recovering  $\gamma a_x^1$ and $\gamma a_x^2[\ell] + \sigma^2 \ell[\ell]$ is immediate from \eqref{eq:mean_micrograph2} and \eqref{eq:ac2_micrograph2}, and $\gamma a_x^3[\ell_1,\ell_2]$ then follows from
%
\begin{align}
\gamma a_x^3[\ell_1,\ell_2]
= a_y^3[\ell_1,\ell_2] - (a_y^1)^3
- a_y^1 \cdot \big(a_y^2[\ell_1]  
+ a_y^2[\ell_2] + a_y^2[\ell_1-\ell_2]- 3(a_y^1)^2 \big).
\end{align}
Using these expressions, we prove the following in Section~\ref{sec:proof_uniqueness_poisson}.
\begin{proposition} \label{prop:uniqueness_poisson}
Under the homogeneous Poisson model, the signal $x$, and the noise level $\sigma^2$ and the Poisson parameter $\gamma$ are identified uniquely. 
In particular, 
	\begin{align}
\gamma &= L \frac{(\gamma a_y^1) (\gamma a_x^2[1])}
{\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] 
	+ \sum_{\ell=2}^{L-1} \gamma a_x^3[\ell,\ell+1]}, \\
\sigma^2 &= a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\frac{L (a^1_y)^2}{\gamma}
- (2 L - 1) (a_y^1)^2.
\end{align}
\end{proposition}

\subsection{Elemetary limitations of the heterogeneous case}
\label{sec:heterogeneity}

@@

\subsection{Proof of Proposition~\ref{prop:uniqueness_poisson}} \label{sec:proof_uniqueness_poisson}
	Let us assume that $x$ is generic. Indeed, we observe the product:
	%
	\begin{align}
	L^2(\gamma a_x^1) (\gamma a_x^2[1])
	&= \gamma^2\left( \sum_{i} x[i] \right) \left(\sum_{j} x[j]x[j+1] \right)
	\nonumber \\
	&= \gamma^2\sum_{j} \sum_{i} x[i]x[j]x[j+1]
	\nonumber \\
	&=  \gamma^2\sum_{j} \sum_{\ell} x[j+\ell]x[j]x[j+1]
	\nonumber \\
	&=  \gamma^2\sum_{\ell \ge 0} \sum_{j} x[j+\ell]x[j]x[j+1] 
	+ \gamma^2\sum_{\ell < 0} \sum_{j} x[j+\ell]x[j]x[j+1]
	\nonumber \\
	&=  \gamma^2\sum_{\ell=0}^{L-1} a_x^3[1,\ell] 
	+ \gamma^2\sum_{\ell > 0} \sum_{j} x[j-\ell]x[j]x[j+1]
	\nonumber \\
	&=  \gamma^2\sum_{\ell=0}^{L-1} a_x^3[1,\ell] 
	+ \gamma^2\sum_{\ell > 0} \sum_{j} x[j]x[j+\ell]x[j+\ell+1]
	\nonumber \\
	&= \gamma\left\{ \sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell]
	+ \sum_{\ell=1}^{L-2} \gamma a_x^3[\ell,\ell+1]  \right\}.
	\end{align}
	%
	Since we also observe $\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] + \sum_{\ell=1}^{L-2} \gamma a_x^3[\ell,\ell+1]$, we can form the ratio and solve for $\gamma$, and then $\sigma^2$:

	\begin{align}
	\gamma = L \frac{(\gamma a_x^1) (\gamma a_x^2[1])}
	{\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] 
		+ \sum_{\ell=2}^{L-1} \gamma a_x^3[\ell,\ell+1]},
	\end{align}
	%
	\begin{align}
	\sigma^2 = a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\frac{L (a^1_y)^2}{\gamma}
	- (2 L - 1) (a_y^1)^2.
	\end{align}
Once $\gamma$ and $\sigma$ were computed, one can recover $x$ from $a_x^2$ and $a_x^3$  by Proposition~\ref{prop:uniqueness}.

\section*{Acknowledgment} \TODO{To check}
The authors  thank Ayelet Heimowitz, Joe Kileel, Ti-Yen Lan and Amit Moscovich  for helpful discussions..
The research was partially supported by Award Number R01GM090200 from the NIGMS, FA9550-17-1-0291 from AFOSR, Simons Foundation Math+X Investigator Award, and the Moore Foundation Data-Driven Discovery Investigator Award.
NB is partially supported by NSF award DMS-1719558.

>>>>

In what follows, we consider autocorrelations of both the observation $y$ and of the signal occurrences in $y$. As per our discussion of extensions, the signal occurrences may be sampled from a discrete set $\{x_1, \ldots, x_K\}$ (as in~\eqref{eq:heterogeneousdistribution}), or from a more general distribution. Accordingly, we define autocorrelations for a random signal $z$ of length $M$. For our purposes, this will be applied both to signal occurrences (of length $L$) and to $y$ (of length $N$).

For a random signal $z\in \R^M$, the autocorrelation of order $q = 1, 2, \ldots$ is given for any integer shifts $\ell_1, \ldots, \ell_{q-1}$ by
\begin{equation}
a_z^q[\ell_1,\ldots,\ell_{q-1}]   = \E_z\left\{\frac{1}{M} \sum_{i=-\infty}^{\infty} z[i]z[i+\ell_1]\cdots z[i+\ell_{q-1}]\right\},
\label{eq:ac_general}
\end{equation}
where the expectation is taken with respect to the distribution of $z$. Indexing of $z$ out of the range $0, \ldots, M-1$ is zero-padded.
Explicitly, the first-, second- and third-order autocorrelations are given by: 
\begin{align}  
a_z^1 & = \E_z\left\{\frac{1}{M} \sum_{i=0}^{M-1} 
z[i]\right\}, \nonumber\\
a_z^2[\ell] & = \E_z\left\{\frac{1}{M} \sum_{i = \max\{0, -\ell\}}^{M-1 + \min\{0, -\ell\}} z[i]z[i+\ell]\right\}, \label{eq:ac_special} \\
a_z^3[\ell_1,\ell_2] & = \E_z\left\{\frac{1}{M} \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{M-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]\right\}. \nonumber
\end{align}
Since  autocorrelations depend only on the differences between indices, they obey the following symmetries: $$a_z^2[\ell] = a_z^2[-\ell],$$ and
$$a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].
$$
%In what follows, $x$ is the random variable corresponding to signal occurrences in $y$.
% In particular, for the model~\eqref{eq:model}, 
%$x=x_k$ with probability $\pi_k$.
In particular, for
$x$ sampled from $\{ x_1, \ldots, x_K \}$ with probabilities $(\pi_1, \ldots, \pi_K)$ as in~\eqref{eq:heterogeneousdistribution}, the autocorrelations are given in explicit form as:
\begin{align}
	a_x^q  & = \sum_{k=1}^{K} \pi_ka_{x_k}^q.
	\label{eq:mixedautocorr}
\end{align}
Explicit expressions of the autocorrelations for the more involved model of  cryo-EM~\eqref{eq:cryoEM} are given in~\cite{bendory2018toward}.

We are given one observation (one realization) of $y$. Thus, we cannot compute the autocorrelations of $y$ exactly as they involve taking an expectation against the distribution of $y$. However, by the law of large numbers, as $N$ grows to infinity with $\gamma$ remaining constant, the empirical autocorrelations of $y$ almost surely (a.s.) converge to the actual (population) autocorrelations of $y$, that is, % (ergodicity)
\begin{align}
\lim_{N \to \infty, \gamma \textrm{ constant}} \frac{1}{N} \sum_{i=-\infty}^{\infty} y[i]y[i+\ell_1]\cdots y[i+\ell_{q-1}] \aseq a_y^q[\ell_1, \ldots, \ell_{q-1}].
\end{align}
\TODO{Why do we need constant $\gamma$?}
This provides a concrete means of estimating the quantities $a_y^ q$. In the remainder of this section, we relate the observables $a_y^q$  to the unknowns $a_x^q$, first under the well-separated model, then under the Poisson model.

@@

Similarly to the homogeneous well-separated model, the observed autocorrelations under the Poisson model identify $x,\gamma$ and $\sigma^2$ uniquely.
Opposed to Proposition~\ref{prop:gamma_sigma}, these quantities can be computed explicitly. 
In Section~\ref{sec:proof_uniqueness_poisson} we prove the following result:
%
%We first note that $\gamma a_x^1$, $\gamma a_x^2[\ell] + \sigma^2 \ell[\ell]$, and $\gamma a_x^3[\ell_1,\ell_2]$ can be computed directly form the observed autocorrelations~\eqref{eq:mean_micrograph2},~\eqref{eq:ac2_micrograph2} and~\eqref{eq:ac3_micrograph2}.
% Indeed, recovering  $\gamma a_x^1$ and $\gamma a_x^2[\ell] + \sigma^2 \ell[\ell]$ is immediate from \eqref{eq:mean_micrograph2} and \eqref{eq:ac2_micrograph2}, and $\gamma a_x^3[\ell_1,\ell_2]$ then follows from
%%
%\begin{align}
%\gamma a_x^3[\ell_1,\ell_2]
%= a_y^3[\ell_1,\ell_2] - (a_y^1)^3
%- a_y^1 \cdot \big(a_y^2[\ell_1]  
%+ a_y^2[\ell_2] + a_y^2[\ell_1-\ell_2]- 3(a_y^1)^2 \big).
%\end{align}
%Using these expressions, we prove the following in Section~\ref{sec:proof_uniqueness_poisson}.
\begin{proposition} \label{prop:uniqueness_poisson}
Under the homogeneous Poisson model, the signal $x$, the noise level $\sigma^2$ and the Poisson parameter $\gamma$ are identified uniquely. 
In particular, \TODO{Shouldn't it be $L^2$ for $\gamma$?}
	\begin{align}
\gamma &= L \frac{(\gamma a_y^1) (a_y^2[1] - (a_y^1)^2)}
{\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] 
	+ \sum_{\ell=2}^{L-1} \gamma a_x^3[\ell,\ell+1]}, \\
\sigma^2 &= a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\frac{L (a^1_y)^2}{\gamma}
- (2 L - 1) (a_y^1)^2,
\end{align}
where 
\begin{align}
\gamma a_x^3[\ell_1,\ell_2]
&= a_y^3[\ell_1,\ell_2] - (a_y^1)^3
- a_y^1 \cdot \big(a_y^2[\ell_1]  
+ a_y^2[\ell_2] + a_y^2[\ell_1-\ell_2]- 3(a_y^1)^2 \big).
\end{align}

\end{proposition}

\subsection{Elementary limitations of the heterogeneous case}
\label{sec:heterogeneity}

@@

\subsection{Proof of Proposition~\ref{prop:uniqueness_poisson}} \label{sec:proof_uniqueness_poisson}

We first note that $\gamma a_x^1$, $\gamma a_x^2[\ell] + \sigma^2 \delta[\ell]$, and $\gamma a_x^3[\ell_1,\ell_2]$ can be computed directly form the observed autocorrelations~\eqref{eq:mean_micrograph2},~\eqref{eq:ac2_micrograph2} and~\eqref{eq:ac3_micrograph2}.
Indeed, recovering  $\gamma a_x^1$ and $\gamma a_x^2[\ell] + \sigma^2 \ell[\ell]$ is immediate from \eqref{eq:mean_micrograph2} and \eqref{eq:ac2_micrograph2}, and $\gamma a_x^3[\ell_1,\ell_2]$ then follows from
%
\begin{align}
\gamma a_x^3[\ell_1,\ell_2]
= a_y^3[\ell_1,\ell_2] - (a_y^1)^3
- a_y^1 \cdot \big(a_y^2[\ell_1]  
+ a_y^2[\ell_2] + a_y^2[\ell_1-\ell_2]- 3(a_y^1)^2 \big).
\end{align}

	Let us assume that $x$ is generic. Indeed, we observe the product:
	%
	\begin{align}
	L^2(\gamma a_x^1) (\gamma a_x^2[1])
	&= \gamma^2\left( \sum_{i} x[i] \right) \left(\sum_{j} x[j]x[j+1] \right)
	\nonumber \\
	&= \gamma^2\sum_{j} \sum_{i} x[i]x[j]x[j+1]
	\nonumber \\
	&=  \gamma^2\sum_{j} \sum_{\ell} x[j+\ell]x[j]x[j+1]
	\nonumber \\
	&=  \gamma^2\sum_{\ell \ge 0} \sum_{j} x[j+\ell]x[j]x[j+1] 
	+ \gamma^2\sum_{\ell < 0} \sum_{j} x[j+\ell]x[j]x[j+1]
	\nonumber \\
	&=  \gamma^2\sum_{\ell=0}^{L-1} a_x^3[1,\ell] 
	+ \gamma^2\sum_{\ell > 0} \sum_{j} x[j-\ell]x[j]x[j+1]
	\nonumber \\
	&=  \gamma^2\sum_{\ell=0}^{L-1} a_x^3[1,\ell] 
	+ \gamma^2\sum_{\ell > 0} \sum_{j} x[j]x[j+\ell]x[j+\ell+1]
	\nonumber \\
	&= \gamma\left( \sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell]
	+ \sum_{\ell=1}^{L-2} \gamma a_x^3[\ell,\ell+1]  \right).
	\end{align}
	%
	Since we also observe $\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] + \sum_{\ell=1}^{L-2} \gamma a_x^3[\ell,\ell+1]$, we can form the ratio and solve for $\gamma$.

	\begin{align}
	\gamma = L \frac{(\gamma a_x^1) (\gamma a_x^2[1])}
	{\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] 
		+ \sum_{\ell=2}^{L-1} \gamma a_x^3[\ell,\ell+1]},
	\end{align}
	%
	Then, similarly to Section~\ref{sec:proof_prop_gamma}, one can solve for $\sigma^2$:
	\begin{align}
	\sigma^2 = a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\frac{L (a^1_y)^2}{\gamma}
	- (2 L - 1) (a_y^1)^2.
	\end{align}
Once $\gamma$ and $\sigma$ were computed, one can recover $x$ from $a_x^2$ and $a_x^3$  by Proposition~\ref{prop:uniqueness}.

\section*{Acknowledgment} \TODO{To check}
The authors  thank Ayelet Heimowitz, Joe Kileel, Ti-Yen Lan and Amit Moscovich  for helpful discussions.
The research was partially supported by Award Number R01GM090200 from the NIGMS, FA9550-17-1-0291 from AFOSR, Simons Foundation Math+X Investigator Award, and the Moore Foundation Data-Driven Discovery Investigator Award.
NB is partially supported by NSF award DMS-1719558.
